\textls[-0]{\thetitle}

### 2023-03-13 metatheory-condensed.tex
comments

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent product types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.
For each derivation (of each judgment) we introduce judgmental equality
(of the form \goodTerm{\Gm}{t_1 \equiv t_2}{T})
indicating convertibility.

We use Tarski-style universes \cite{hofmann1997syntax},
which, unlike Russell-style ones,
are not inhabited by types directly but rather by the \emph{codes} of types
and arguably behave better \cite{luo2012notes}.
Term $\codety{T}$ encodes type~$T$ (\ruleref{tm/c}) and thus inhabits universe~$\cU$.
Type $\El{t}$ decodes term~$t$ (\ruleref{ty/el}).
%\footnote{By contrast,  Russel-style universes are inhabited by types directly.  Tarski-style universes are ``semantically more fundamental'' \cite{luo2012notes} and friendlier to our meta-theoretical reasoning.}
There is an infinite hierarchy of universes;
we omit universe levels in the presentation for conciseness.
%\rA{Why is W(t) a term and not a type?}\YZreply{I think the reviewer is wondering what is gained by encoding and decoding types, compared to types directly inhabiting universes.}

% We use intrinsically typed syntax \cite{church1940}---terms and types are well-formed by construction.
For conciseness, typing rules omit obvious premises required for well-formedness.
For example, \ruleref{tm/lam} implicitly requires that the context be well-typed
($\goodCtx{\Gm}{}$).

@@

W-types~\cite{martin1984intuitionistic} are a succinct way to model
inductive types in MLTT.
Together with the identity type $\TyId{\cDot}{\cDot}$, they can express a
whole host of inductive types~\cite{hugunin2021not}.
%
Our enhanced W-type is decided by a W-type signature $\goodWSig{\Gm}{\wsig}{n}$ 
composed of a list (\ruleref{wsig/empty} and \ruleref{wsig/add}) of $n$ pairs of 
types $\{(\wsigproj{i}{1}{\wsig}, \wsigproj{i}{2}{\wsig})\}_i$,
modeling an inductive type of $n$ constructors. 
%
Rule~\ruleref{tm/wsup} states that the $i$-th pair $\goodType{\Gm}{\wsigproj{i}{1}{\wsig}}{}$ and 
$\goodType{\Gm,\wsigproj{i}{1}{\wsig}}{\wsigproj{i}{2}{\wsig}}{}$ models the $i$-th constructor---given 
two arguments $\goodTerm{\Gm}{t_1}{\wsigproj{i}{1}{\wsig}}$ and
$\goodTerm{\Gm, \mathit{self} : \nsub{\wsigproj{i}{2}{\wsig}}{x}{t_1}}{t_2}{\El{\wcode{\wsig}}}$,
we can construct a term $\Wsup{i}{\wsig}{t_1}{\mathit{self}.t_2}$ of the W-type. $\wcode{\wsig}$ gives the code of the
W-type (\ruleref{tm/w}).

>>>>

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent pair types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.
For each derivation (of each judgment)\YZ{I don't get what 'each derivation' or 'each judgment' means}
we introduce judgmental equality (of the form \goodTerm{\Gm}{t_1 \equiv t_2}{T})
indicating convertibility.

We use Tarski-style universes \cite{hofmann1997syntax},
which, unlike Russell-style ones,
are not inhabited by types directly but rather by the \emph{codes} of types
and arguably behave better \cite{luo2012notes}.
Term $\codety{T}$ encodes type~$T$ (\ruleref{tm/c}) and thus inhabits universe~$\cU$.
Type $\El{t}$ decodes term~$t$ (\ruleref{ty/el}).
%\footnote{By contrast,  Russel-style universes are inhabited by types directly.  Tarski-style universes are ``semantically more fundamental'' \cite{luo2012notes} and friendlier to our meta-theoretical reasoning.}
There is an infinite hierarchy of universes;
we omit universe levels in the presentation to avoid clutter.
%\rA{Why is W(t) a term and not a type?}\YZreply{I think the reviewer is wondering what is gained by encoding and decoding types, compared to types directly inhabiting universes.}

% We use intrinsically typed syntax \cite{church1940}---terms and types are well-formed by construction.
For concision, typing rules omit obvious premises required for well-formedness.
For example, \ruleref{tm/lam} implicitly requires $\goodCtx{\Gm}{}$
(i.e., that the context be well-typed).

@@

W-types~\cite{martin1984intuitionistic} are a succinct way to model
inductive types in MLTT.
Together with the identity type, they can express a
whole host of inductive types~\cite{hugunin2021not}.
%
Our enhanced W-type is decided by a W-type signature $\goodWSig{\Gm}{\wsig}{n}$ 
composed of a list (\ruleref{wsig/empty} and \ruleref{wsig/add}) of $n$ pairs of 
types $\{(\wsigproj{i}{1}{\wsig}, \wsigproj{i}{2}{\wsig})\}_i$,
modeling an inductive type of $n$ constructors. 
%
Rule~\ruleref{tm/wsup} states that the $i$-th pair $\goodType{\Gm}{\wsigproj{i}{1}{\wsig}}{}$ and 
$\goodType{\Gm,\wsigproj{i}{1}{\wsig}}{\wsigproj{i}{2}{\wsig}}{}$\YZ{this is still in de Brujin}
models the $i$-th constructor---given
two arguments $\goodTerm{\Gm}{t_1}{\wsigproj{i}{1}{\wsig}}$ and
$\goodTerm{\Gm, \mathit{self} : \nsub{\wsigproj{i}{2}{\wsig}}{x}{t_1}}{t_2}{\El{\wcode{\wsig}}}$,
we can construct a term $\Wsup{i}{\wsig}{t_1}{\mathit{self}.t_2}$ of the W-type. $\wcode{\wsig}$ gives the code of the
W-type (\ruleref{tm/w}).\YZ{What's the rationale behind naming the free variable 'self'?}
